---
title: "Lab 10"
date: "3/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<style type='text/css'>
  body{
  font-size: 14pt;
  }
  pre {
  font-size: 12pt;
  }
</style>

```{r call libraries, message=F}
library(car)
library(tidyverse)
library(ggplot2)
library(ez)
library(lme4)
library(emmeans)
library(MASS)
if(!require(nlme)) install.packages('nlme')
if(!require(bruceR)) install.packages('bruceR')
```

# Analyze Student Calories dataset

Students were asked to document their daily caloric intake once a month for six months.  Students were divided into three groups with each receiving instruction in nutrition education using one of three curricula.  

Goal: To test for differences in the calories per day between cirricula. 


Read in data:
```{r read in student data}

student.df <- readRDS('Student_Calories.rds')

head(student.df)
```

```{r convert factors}

student.df <- student.df %>% 
  mutate(Instruction=factor(Instruction),
         Student=factor(Student),
         Month=factor(Month))
```


```{r descriptives statistics}

summary(student.df)

```

```{r density plot}
ggplot(student.df, aes(x=Calories.per.day)) + 
  geom_density(fill='blue', color='black', alpha=0.3) +
  labs(x='Calories Per Day', y='Density', title='Student Calories Per Day') + 
  theme_classic() + 
  theme(plot.title=element_text(hjust=0.5, size=14))
```
```{r test normality}
shapiro.test(student.df$Calories.per.day)
```

```{r two way repeated measures anova}

student.model <- MANOVA(student.df, subID='Student', dv=c('Calories.per.day'),
                        between=c('Instruction'), with='Month')
```
```{r}
student.model.2 <- MANOVA(student.df, subID='Student', dv=c('Calories.per.day'),
                        between=c('Instruction'), with='Month', sph.correction='GG')
```

```{r extract means}
month_instruction_means <- student.df %>% 
  group_by(Month, Instruction) %>% 
  summarize(mu = mean(Calories.per.day),
            mu_se = sd(Calories.per.day)/n())
```

```{r plot interaction}
ggplot(month_instruction_means, aes(x=Month, y=mu, color=Instruction)) + 
  geom_errorbar(aes(ymin=mu-mu_se, ymax=mu+mu_se), show.legend=F, width=.1, color='black') +   geom_point(size=2) + 
  geom_line(aes(group=Instruction), show.legend=F, size=1.2) + 
  labs(y='Average Calories Per Day') + 
  theme_minimal()
```

```{r simple effects}
month.1 <- subset(student.df, Month=='1')
month.2 <- subset(student.df, Month=='2')
month.3 <- subset(student.df, Month=='3')
month.4 <- subset(student.df, Month=='4')
month.5 <- subset(student.df, Month=='5')
month.6 <- subset(student.df, Month=='6')

instruction.A <- subset(student.df, Instruction=='Curriculum A')
instruction.B <- subset(student.df, Instruction=='Curriculum B')
instruction.C <- subset(student.df, Instruction=='Curriculum C')
```


```{r one-way repeated anovas, message=F}
ezANOVA(month.1, dv=Calories.per.day, wid=Student, between=Instruction)$ANOVA
ezANOVA(month.2, dv=Calories.per.day, wid=Student, between=Instruction)$ANOVA
ezANOVA(month.3, dv=Calories.per.day, wid=Student, between=Instruction)$ANOVA
ezANOVA(month.4, dv=Calories.per.day, wid=Student, between=Instruction)$ANOVA
ezANOVA(month.5, dv=Calories.per.day, wid=Student, between=Instruction)$ANOVA
ezANOVA(month.6, dv=Calories.per.day, wid=Student, between=Instruction)$ANOVA
```

```{r instruction simple effects, message=F}
ezANOVA(instruction.A, dv=Calories.per.day, wid=Student, within=Month)$ANOVA

ezANOVA(instruction.B, dv=Calories.per.day, wid=Student, within=Month)$ANOVA

ezANOVA(instruction.C, dv=Calories.per.day, wid=Student, within=Month)$ANOVA
```

```{r simple contrasts}
student.margins <- emmeans(student.model.2, ~Instruction*Month)

pairs(student.margins, simple='each', adjust='tuk')
```


# Analyze Machines data

A company wanted to replace the machines used to make a certain component. Three different brands of machines were available, so the investigators designed an experiment to evaluate the productivity of the machines when operated by the company's own personnel. 6 `workers` participated in the experiment, each of whom was to operate each machine 3 different times. The response is an overall productivity `score`.   
Goal: to investigate effects of `machine` and `worker` on the score. There are two factors: `machine` and `worker`.  

`Machines` dataset is from the `nlme` package.

Read in data:

```{r read in Machines data}
machine.df <- Machines
head(machine.df)
```

```{r}
summary(machine.df)
```
```{r}
ggplot(machine.df, aes(x=score)) + 
  geom_density(fill='green', alpha=0.3) + 
  labs(x='Productivity Score', y='Density', title='Worker Productivity Scores') + 
  theme_classic() + 
  theme(plot.title=element_text(hjust=0.5, size=14))
```

```{r}
shapiro.test(machine.df$score)
```

```{r}
machine.model <- MANOVA(machine.df, subID='Worker', dv='score', with='Machine')
```

```{r}
machine_means <- as.data.frame(machine.margins)

colnames(machine_means) <- c('Machine','mu','SE','df','lower.CL','upper.CL')
```

```{r}
ggplot(machine_means, aes(x=Machine, y=mu)) + 
  geom_errorbar(aes(ymin=mu-SE, ymax=mu+SE), show.legend=F, width=.1, color='black') +   geom_point(size=3, color='green') + 
  labs(y='Average Productivity Score') + 
  theme_minimal()
```
```{r}
machine.margins <- emmeans(machine.model, ~ Machine)

pairs(machine.margins, adjust='tuk')
```

# Analyze Michelson data 

Dataset contains measurements of the speed of light in air, made between June 5th and July 2nd, 1879. It consists of 5 experiments (`Expt`) each consisting of 20 consecutive `runs`. The response is the `speed` of light in km/s, less 299000. THe currently accepted value, on this scale of measurement, is 734.5.

```{r read in Michelson data}
michelson.df <- michelson
head(michelson.df)
```




