---
title: "Lab 10"
date: "3/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<style type='text/css'>
  body{
  font-size: 14pt;
  }
  pre {
  font-size: 12pt;
  }
</style>

```{r call libraries, message=F}
library(car)
library(tidyverse)
library(ggplot2)
library(ez)

library(MASS)
if(!require(nlme)) install.packages("nlme")
```

# Analyze Student Calories dataset

Students were asked to document their daily caloric intake once a month for six months.  Students were divided into three groups with each receiving instruction in nutrition education using one of three curricula.  

Goal: To test for differences in the calories per day between cirricula. 


Read in data:
```{r read in student data}

student.df <- readRDS('Student_Calories.rds')

head(student.df)
```

# Analyze Machines data

A company wanted to replace the machines used to make a certain component. Three different brands of machines were available, so the investigators designed an experiment to evaluate the productivity of the machines when operated by the company’s own personnel. 6 `workers` participated in the experiment, each of whom was to operate each machine 3 different times. The response is an overall productivity `score`.   
Goal: to investigate effects of `machine` and `worker` on the score. There are two factors: `machine` and `worker`.  

`Machines` dataset is from the `nlme` package.

Read in data:

```{r read in Machines data}
machine.df <- Machines
head(machine.df)
```
# Analyze Michelson data 

```{r read in Michelson data}
michelson.df <- michelson
head(michelson.df)
```




