---
title: "Lab 2"
output:
  html_document:
    df_print: paged
---

# R basics

## Variables

### Character
```{r}
my_character <- 'hello'
```

### Numeric
```{r}
my_numeric <- 3
```

### Logical
```{r}
my_logical <- c(T,F)
```

Check types 
```{r}
typeof(my_character)
typeof(my_numeric)
typeof(my_logical)
```


## Dataframes
### Creating a data frame
```{r}
my_df <- data.frame('a'=c(1,2,3),
                    'b'=c('Jack','Jill','Tony'),
                    'c'=c(T,F,T))


```
### Indexing

+ Single Column
```{r}
my_df$a
```

+ Single Row
```{r}
my_df[1,]
```

+ Multiple Columns and Rows
```{r}
my_df[c(1,2), c(1,3)]
```
## For Loop
```{r}
my_vector <- c(1,2,3,4)

for (value in my_vector){
  new_value <- value+2
  
  print(new_value)
}
```

## Function
```{r}
my.first_function <- function(x){
  f_x <- 3*x + 2
  
  return(f_x)
}


my.first_function(5)
```

## Apply Function

R has a built in special function known as `(l/s/t/m)apply`. This allows you to apply another function to a vector, list, or data frame of values. For instance, we can apply `my.first_function` to a vector:
```{r}
my_vector.2 <- c(5,-2,3,12,1,0)


sapply(my_vector.2, my.first_function)
```
Variations of the apply function differ in how they output results. `lapply` outputs results in a list, while `sapply` outputs them in a vector. Keep in mind that you can also use a for loop to accomplish the same procedure as the apply function, as well as combine the two together.

# Matrix Alegbra in R

```{r}
x <- 5 # create a scalar

v <- c(1,2,3,4) # create a vector

A <- matrix(c(3, 2, 1,
                     7, 5, 4,
                     8, 3, 6), ncol=3, byrow=T) # create a matrix

v
A
```

When creating a matrix, you need to specify either the number of rows or columns that are in the matrix. Here we specify that there are 3 columns using `ncol=3`.
Notice the `byrow=T` flag in the matrix command. By default, R creates matrices by columns. This can be a little confusing, especially if you are used to creating matrices in other languages such as MATLAB. Setting the flag byrow equal to True overrides the default behavior.

Lets start doing some algebra!

```{r}
x + v # scalar plus a vector
```

```{r}
m <- c(12,-10,0,2) # create another vector

m + v # vector addition
```

```{r}
B <- matrix(c(5, 10, 9,
              -3, 0, 0,
              1, 18, 6,
              0, -2, 14), ncol=3, byrow=T) # create a 4 x 3 matrix

v + B
```

```{r}
A + A
```

To multiply matrices and vectors, you have to use the notation %*%
```{r}
B %*% A
```

Use the `solve()` function to find the inverse of matrices, and to divide one matrix by a vector or another matrix.
```{r}
solve(A)
```
```{r error=T}
solve(B)
```
If the matrix does not have an inverse (i.e. it is not square or is *singular*), then the `solve` function will not work.

```{r}
C <- matrix(c(-5, 8, 3,
              2, 0, 4,
              1, 6, 0), ncol=3, byrow=T) 

solve(C,A)
```
```{r}
b <- c(5,-2,1)

solve(A,b)
```

Although only square matrices have a true inverse, we can use a computational tool such as R to calculate *psuedo* inverses. Do do so, we will need to use the *MASS* package.
```{r}
library(MASS)
ginv(B)
```

