---
title: "Lab 6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style type='text/css'>
  body{
  font-size: 14pt;
  }
  pre {
  font-size: 12pt;
  }
</style>

```{r call librarys, message=F}
library(car)
library(tidyverse)
```

In today's lab we are going to conduct a one-way ANOVA. First we'll do it by hand, then using a model created with `lm()`, and lastly with the function `aov()`. Note, that `aov()` is a wrapper for `lm()`.


## One-way ANOVA

As a quick review, we conduct a one-way ANOVA when we want to compare more than two group means. By comparing the variances of each group, we can circumvent the need for conducting multiple *t*-tests and mitigate the chance of a Type I error from becoming too high. In ANOVA, we compute the ratio of between group variance (i.e. $SS_{treatment}$) divided by the amount of within group variance $SS_{error}$. Then, we compare this ratio (i.e. **F-value**) to what is expected under the null hypothesis. 

To compute $SS_{treatment}$ we sum the squared differences of each group mean from the grand mean:

$$
\begin{align}
SS_{treatment} = \sum_{i=1}^{p}\sum_{j=1}^{n_i}(\bar{X}_{i.} - \bar{X}_{..})^2
\end{align}
$$

And to compute $SS_{error}$ you sum the squared difference of each individuals score from their respective group means:   

$$
\begin{align}
SS_{error} = \sum_{i=1}^{p}\sum_{j=1}^{n_i}(\bar{X}_{ij} - \bar{X}_{i.})^2
\end{align}
$$

## By Hand

Lets see this in action by hand. We'll work with the `diet.csv` dataset. As a reminder, this data set contains the following columns:    

+ `Person` ID
+ `gender`
+ `Age`
+ `Height`
+ `pre.weight` - weight at the start of the weight loss program
+ `Diet` - type of diet subject was put on
+ `weight6weeks` - weight at the end of the 6 week program

```{r load & process data}

diet.df <- read.csv('diet.csv')

diet.df <- diet.df %>% 
  rename(final.weight = weight6weeks) %>% # rename weight6weeks column to final.weight
  mutate(weight.loss = final.weight - pre.weight,
         Diet = factor(Diet, labels=c('A','B','C')),
         gender = factor(gender, labels=c('F','M')),
         Person = factor(Person)) %>% 
  drop_na() # drop missing data

head(diet.df, 3)
```

The following assumptions must be met when conducting an ANOVA:

+ Dependent variable is normally distributed
+ Population variance in each group is equal (i.e. Homogeneity of Variances)
+ Observations are independent

Homogeneity of variance can be evaluated using a Levene's Test, which test the null hypothesis that the variance within each group is equal.

```{r Levenes Test}
leveneTest(weight.loss ~ Diet, data=diet.df)
```

A nonsignificant *F* statistic indicates that we fail to reject the null, implying that variance is equal across diet groups.   

Now lets compute the sum of squared error terms by hand: 

```{r SS by hand, message=F}

grand_mean <- mean(diet.df$weight.loss)

# Extract group mean errors
groupMean_errors <- diet.df %>% 
  group_by(Diet) %>% 
  mutate(error_term = (mean(weight.loss) - grand_mean)^2) %>% 
  ungroup()

SS_T <- sum(groupMean_errors$error_term)

residuals <- diet.df %>% 
  group_by(Diet) %>% 
  summarise(error = weight.loss - mean(weight.loss))
  
SS_E <- sum(residuals$error ^ 2)


print(paste('SS Treatment: ', SS_T, 'SS Error: ', SS_E))
```

Now average these sum of squared error terms metrics using the degrees of freedom:

$$
\begin{align}
MST = \frac{SST}{p-1} \\
\\
MSE = \frac{SSE}{np-p}
\end{align}
$$
where $p$ is the number of groups and $n$ is the number of observations in each group. Note, that $np$ is equal to the total number of observations in the whole dataset. So we can subtract that number from the number of groups we have.

```{r MS by hand}

p <- nlevels(diet.df$Diet)
total_n <- nrow(diet.df)

treatment_df <- p-1
residual_df <- total_n-p

MST <- SS_T/treatment_df

MSE <- SS_E/residual_df

```

Next we can compute an F-statistic and compare it to the expected value under the null distribution:

```{r F value }

F.ratio <- MST/MSE

# pf gives us the probability of a value in an F-distribution with specified degrees of freedom
F.p <- pf(F.ratio,treatment_df,residual_df,lower.tail=F)
```


Now lets put it in our very own ANOVA table: 

```{r Custom ANOVA Table}

customAnova.table <- data.frame(Df=c(treatment_df, residual_df),
                           'Sum Sq' = c(SS_T, SS_E),
                           'Mean Sq' = c(MST, MSE),
                           'F value' = c(F.ratio, NA),
                           'Pr(>F)' = c(F.p, NA),
                           row.names = c('Diet','Residuals'))
round(customAnova.table, 3)
```

Congrats! You've now conducted an ANOVA by hand. But it was a lot of code and obviously there are much easier ways to do this.

## Using `lm` (i.e. GLM)

Now lets do it using the `lm` function. We'll impose the sum to zero constraint to follow tradition, but this is actually unnecessary:

```{r lm model}
model.1 <- lm(weight.loss ~ Diet, data=diet.df, contrasts = list(Diet = contr.sum))
```

Remember, when we multiple the estimated $\beta$ coefficients by the design matrix, we get the means of each group. Here we'll use a toy design matrix with a single individual from each group:   

```{r lm coefs}

betas <- coef(model.1)
X <- model.matrix(~Diet, expand.grid(Diet=levels(diet.df$Diet)), contrast=list(Diet=contr.sum))

X %*% betas
```
Using the estimated coefficients in this fashion is also known as making "predictions". The design matrix can be treated as some new data that the model has never seen. These predictions can be made using the `predict()` function. If no new data is passed in as an input, then the predict function just uses the model's design matrix (i.e., already observed values). So we can use the `predict()` function to get the group means for each of our observations. Note, predictions are formally expressed as: 

$$
\begin{aligned}
\hat{y} = X\hat{\beta}
\end{aligned}
$$
where $\hat{y}$ is the predicted value. Lets get those group means and subtract them from the grand mean.

```{r predictions}
lm.SST <- sum((predict(model.1) - grand_mean)^2)

lm.SST
SS_T
```
As you can see, the computed $SS_{treatment}$ with `lm` is exactly what we got by hand. Next, we need the residual sum of squared error. This is already stored in the model:

```{r lm residuals}

lm.SSE <- sum((model.1$residuals)^2)

lm.SSE
SS_E
```
Remember, the residuals are $\epsilon = y - \hat{y}$, where $\hat{y}$ is the predicted value (i.e. group means).  

To get the full ANOVA table, we would just follow the same steps that we did by hand before.

## Working with `aov()`


Now we can use the function `aov()` to derive the ANOVA table. The function uses a similar formula to that which we used for `lm`.   

```{r}

model.2 <- aov(weight.loss ~ Diet, data=diet.df)
  
summary(model.2)  
```
Boom! Exactly what we calculated by hand and with the `lm` model. Now its your turn! Conduct a One-way ANOVA by hand with the `iris` dataset from the *car* package, comparing the mean `Petal.Length` between each `Species` of iris. 

```{r load iris data}
data(iris)

head(iris)
```





